---
title: "Body Mass & Physical Demands"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

players <- read_csv("data/players.csv", show_col_types = FALSE) |>
  mutate(bmi = round((weight / (height_inches^2)) * 703, 1))
```

## How Positional Roles Shape Body Density

In baseball, sheer height only tells part of the story. Body Mass Index (BMI)—calculated from an athlete's weight relative to height—illustrates the structural density required by different fielding responsibilities.

```{r}
#| fig-width: 8
#| fig-height: 5

players |>
  filter(!is.na(bmi), !is.na(position)) |>
  mutate(position = fct_reorder(position, bmi, .fun = median)) |>
  ggplot(aes(x = bmi, y = position, fill = position)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(
    title = "BMI Distribution by Defensive Position (2024 MLB)",
    x = "Body Mass Index (BMI)",
    y = "Position",
    caption = "Source: 2024 MLB Player Profiles"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

First basemen and catchers exhibit the highest median BMIs across the league, reflecting the muscular bulk and leverage needed for power hitting and absorbing home-plate contact. In contrast, shortstops and second basemen maintain leaner frames to maximize lateral mobility and quick reaction times.
