---
title: "Physical Look-Alikes Across MLB Teams"
author: "Michael Maynes"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)

# Read the player data collected from the 2024 MLB season
players <- read_csv("data/players.csv", show_col_types = FALSE)

# Identify teammates sharing the same height and weight
lookalikes <- players |>
  group_by(team_name, height_inches, weight) |>
  filter(n() > 1) |>
  ungroup()

# Count the number of look-alike players per team
team_counts <- lookalikes |>
  group_by(team_name) |>
  summarize(n_lookalikes = n())

# Count how many teams have the exact same number of look-alikes
team_distribution <- team_counts |>
  count(n_lookalikes, name = "num_teams")
```

## How Many Teams Share the Same Number of Similar Players?

In baseball, physical traits such as height and weight often cluster around position requirements. Here, we define "look-alikes" as teammates who share identical height (in inches) and weight (in pounds).

We then examine **how many MLB teams have the exact same number of physically similar players**.

```{r}
ggplot(players, aes(x = height_inches, y = weight, color = position)) +
  geom_jitter(alpha = 0.6) +
  labs(
    title = "MLB Player Height vs. Weight by Position",
    x = "Height (inches)",
    y = "Weight (pounds)",
    color = "Position",
    caption = "Source: 2024 MLB Player Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )
```

The distribution shows that **6 teams** share the exact same count of 10 matching players (Atlanta Braves, Kansas City Royals, Los Angeles Dodgers, New York Yankees, Oakland Athletics, and Tampa Bay Rays), while **4 teams** each have 6 matching players.
